\textbf{The Mint Scale: A Fresh Validation of the Multimodal Interoception Questionnaire and Comparison to the MAIA, BPQ and IAS}

### 2025-08-19 paper/manuscript.tex
manuscript pdf

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Giulia
Poreiro}{0000-0002-2343-5109}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, opacityback=0, colback=white, breakable, toprule=.15mm, arc=.35mm, left=2mm, rightrule=.15mm, bottomrule=.15mm, colframe=quarto-callout-note-color-frame]

This preprint is a non-peer-reviewed work from the
\href{https://realitybending.github.io/}{\textbf{Reality Bending Lab}}.
\includegraphics[width=0.2\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/ReBeL_LogoOnly_hu114.png}

@@

Interoception - the sensing and interpretation of the body's internal
physiological signals - plays a fundamental role in processes ranging
from homeostatic regulation to emotion, decision-making, and mental
health (\citeproc{ref-craig2002how}{Craig, 2002};
\citeproc{ref-critchley2017interoception}{Critchley \& Garfinkel,
2017}). Despite its growing recognition, the field has been hindered by
inconsistent definitions and fragmented measurement approaches
(\citeproc{ref-desmedt2022measures}{Desmedt et al., 2022},
\citeproc{ref-desmedt2025discrepancies}{2025};
\citeproc{ref-khalsa2018interoception}{Khalsa et al., 2018}). Available
tools include ``objective'' physiological tasks (e.g., heartbeat
counting or tracking) and ``subjective'' self-reports, and encompass
both ``explicit'' paradigms that direct participants' attention to
internal sensations and ``implicit'' paradigms that do not (e.g.,
heartbeat evoked potentials recorded during rest). These assessments are
often targeted at only one of several interoceptive modalities (e.g.,
cardioception, respiroception, gastroception), and are mapped onto
theoretical dimensions (e.g., accuracy, sensitivity, awareness,
\citeproc{ref-garfinkel2015knowing}{Garfinkel et al., 2015}) within a
rapidly evolving theoretical landscape. However, no consensus exists on
a gold-standard measure witheach approach having distinct limitations.
For example, physiological tasks often require specialised equipment and
expertise, are time-consuming, and are challenging to administer at
scale, whereas questionnaires are constrained by their inherent
subjectivity. Crucially, the psychometric quality of many of these
measures is questionable (e.g., see criticism about the heartbeat
counting task, \citeproc{ref-zamariola2018interoceptive}{Zamariola et
al., 2018}), yielding low inter-task correlations
(\citeproc{ref-murphy2018classifying}{Murphy et al., 2018}) and raising
important concerns regarding their reliability and validity (e.g.,
measures presented as measuring one theoretical dimension but likely
measuring something else). This lack of conceptual clarity and
measurement convergence impedes scientific progress in comparing,
replicating, and interpreting interoception-related findings,
emphasising the need to develop improved tools that align with modern
psychometric standards.

@@

These differing philosophies have practical consequences, leading to
poor convergence between scales that are assumed to measure the same
construct (\citeproc{ref-vig2022questionnaires}{Vig et al., 2022}) - a
classic example of the ``jingle fallacy'' where the shared label of
``interoception'' can mask underlying measurand differences, hindering
replication and theoretical integration. In acknowledgement of this
issue, efforts have been made to align subjective measures with
theoretical models that outline different facets of interoception (e.g.,
accuracy, attention, deficits). However, focusing on developing measures
that are theory-driven may however be premature when theoretical models
are in flux, and when a topographic mapping to self-report dimensions is
not guaranteed. For example, interoceptive distinctions that are
relevant at a neuroanatomical or cognitive level might not necessarily
map onto the structure of subjective experiences. More fundamentally,
existing interoception questionnaires typically fail to consider the
full range of interoceptive modalities and the context in which
interoceptive experiences occur. Most existing questionnaires focus on a
narrow range of bodily modalities, with a heavy bias towards
cardio-respiratory sensations at the expense of others like gastric,
urogenital, or thermoregulatory signals. Perhaps most critically,
existing measures ask participants to make general assessments across
most/all contexts, which does not capture the potential homeostatic
relevance of attending or detecting internal signals. Without a
contextual reference point, there is likely to be substantial
variability in how participants interpret and respond to items. For
example, a general item about ``feeling your heart beat'' is ambiguous
since its salience, frequency, experience and interpretation would
differ depending on the context, e.g., on whether it occurs during
physical exercise, a moment of anxiety, or quiet rest (e.g., someone who
exercises regularly, or is anxious, might more easily recall an instance
of occurrence of feeling their heart beating and respond based on that
contextual saliency).

@@

To further reduce and balance the remaining items, we collectively
decided on removing the lower-level clusters \emph{Sexual Arousal
Awareness} (too general and overlapping with the other more specific
sex-related items), \emph{Signalling} (which items started with ``when
something important is happening in my life'', which meaning we deemed
too much open to interpretation), and \emph{Sexual Arousal Changes} (low
fit with the other modality-focused clusters of its group). The final
set included 45 items.

@@

Next, we tested the predictive power of the Mint scale relative to other
interoception questionnaires. We will assessed the importance of each
interoceptive dimension (from all the scales) as a unique predictor by
fitting regression models (linear for continuous measures - e.g.,
depression score - and logistic for binary variables - e.g., presence
\emph{vs.} absence of mood disorders) to predict different outcomes, and
compare the best between the 4 interoception scales (based on the R2).
We then evaluated the predictive performance of each scale as a whole by
comparing regression models with all the dimensions entered as
predictors (note that the IAS and BPQ only have one total score
variable). We assessed the models based on their total explained
variance (R2), as well as on the Bayesian Information Criterion (BIC),
which penalizes for predictor number, thus offering a balance between
model parsimony and predictive power. For the logistic models, we
quantified the discriminative power by computing the Area Under the
Curve (AUC) of the Receiver Operating Characteristic (ROC) curve, which
assesses the model's discriminative power (the combination of
sensitivity and specificity).

@@

The correlation matrix of the Mint dimensions revealed an interesting
and rich tapestry of relationships (see Figure~\ref{fig-four}), with
contrasting patterns of associations with dimensions from the same
group. For instance, \emph{Derm} and \emph{Olfa}, despite being
positively correlated with the other dimensions from the \emph{Deficit}
cluster, were not negatively correlated to \emph{Awareness} and its
dimensions. Similarly, \emph{Sati}, unlike the other \emph{Deficit}
dimensions, was not positively correlated to \emph{Visceroception}. This
complex structure suggests that the dimensions indeed capture distinct
aspects of interoception, and that the metaclusters, rather than being
simple aggregates of rather-redundant elements, might actually capture
unique combinations (greater than the sum of their parts). It also
provides evidence against a simplistic adaptive \emph{vs.} maladaptive
dichotomy, as \emph{Deficit} was not negatively bit positively
correlated with \emph{Visceroception}. Interestingly, while
\emph{Awareness} was also positively correlated with
\emph{Visceroception}, it yielded an insignificant negative correlation
with \emph{Deficit} (this finding was also aligned with the hierarchical
clustering results, showing a greater proximity between \emph{Deficit}
and \emph{Visceroception} than with \emph{Awareness}), underlining again
the complex web of relationships captured by the Mint.

>>>>

\authornote{\par{\addORCIDlink{Dominique
Makowski}{0000-0001-5375-9967}}\par{\addORCIDlink{Ana
Neves}{0009-0006-0020-7599}}\par{\addORCIDlink{Giulia
Poreiro}{0000-0002-2343-5109}} 
\par{ }
\par{     \begin{tcolorbox}[enhanced jigsaw, breakable, colframe=quarto-callout-note-color-frame, leftrule=.75mm, arc=.35mm, colback=white, toprule=.15mm, bottomrule=.15mm, rightrule=.15mm, opacityback=0, left=2mm]

This preprint is a non-peer-reviewed work from the
\href{https://realitybending.github.io/}{\textbf{Reality Bending Lab}}.
\begin{center}
\includegraphics[width=0.2\linewidth,height=\textheight,keepaspectratio]{manuscript_files/mediabag/ReBeL_LogoOnly_hu114.png}
\end{center}

@@

Interoception - the sensing and interpretation of the body's internal
physiological signals - plays a fundamental role in processes ranging
from homeostatic regulation to emotion, decision-making, and mental
health (\citeproc{ref-craig2002how}{Craig, 2002};
\citeproc{ref-critchley2017interoception}{Critchley \& Garfinkel,
2017}). Despite its growing recognition, the field has been hindered by
inconsistent definitions and fragmented measurement approaches
(\citeproc{ref-desmedt2022measures}{Desmedt et al., 2022},
\citeproc{ref-desmedt2025discrepancies}{2025};
\citeproc{ref-khalsa2018interoception}{Khalsa et al., 2018}). Available
tools include ``objective'' physiological tasks (e.g., heartbeat
counting or tracking) and ``subjective'' self-reports, and encompass
both ``explicit'' paradigms that direct participants' attention to
internal sensations and ``implicit'' paradigms that do not (e.g.,
heartbeat evoked potentials recorded during rest). These assessments are
often targeted at only one of several interoceptive modalities (e.g.,
cardioception, respiroception, gastroception), and are mapped onto
theoretical dimensions (e.g., accuracy, sensitivity, awareness,
\citeproc{ref-garfinkel2015knowing}{Garfinkel et al., 2015}) within a
rapidly evolving theoretical landscape. However, no consensus exists on
a gold-standard measure with each approach having distinct limitations.
For example, physiological tasks often require specialised equipment and
expertise, are time-consuming, and are challenging to administer at
scale, whereas questionnaires are constrained by their inherent
subjectivity. Crucially, the psychometric quality of many of these
measures is questionable (e.g., see criticism about the heartbeat
counting task, \citeproc{ref-zamariola2018interoceptive}{Zamariola et
al., 2018}), yielding low inter-task correlations
(\citeproc{ref-murphy2018classifying}{Murphy et al., 2018}) and raising
important concerns regarding their reliability and validity (e.g.,
measures presented as measuring one theoretical dimension but likely
measuring something else). This lack of conceptual clarity and
measurement convergence impedes scientific progress in comparing,
replicating, and interpreting interoception-related findings,
emphasising the need to develop improved tools that align with modern
psychometric standards.

@@

These differing philosophies have practical consequences, leading to
poor convergence between scales that are assumed to measure the same
construct (\citeproc{ref-vig2022questionnaires}{Vig et al., 2022}) - a
classic example of the ``jingle fallacy'' where the shared label of
``interoception'' can mask underlying measures and differences,
hindering replication and theoretical integration. In acknowledgement of
this issue, efforts have been made to align subjective measures with
theoretical models that outline different facets of interoception (e.g.,
accuracy, attention, deficits). However, focusing on developing measures
that are theory-driven may however be premature when theoretical models
are in flux, and when a topographic mapping to self-report dimensions is
not guaranteed. For example, interoceptive distinctions that are
relevant at a neuroanatomical or cognitive level might not necessarily
map onto the structure of subjective experiences. More fundamentally,
existing interoception questionnaires typically fail to consider the
full range of interoceptive modalities and the context in which
interoceptive experiences occur. Most existing questionnaires focus on a
narrow range of bodily modalities, with a heavy bias towards
cardio-respiratory sensations at the expense of others like gastric,
urogenital, or thermoregulatory signals. Perhaps most critically,
existing measures ask participants to make general assessments across
most/all contexts, which does not capture the potential homeostatic
relevance of attending or detecting internal signals. Without a
contextual reference point, there is likely to be substantial
variability in how participants interpret and respond to items. For
example, a general item about ``feeling your heart beat'' is ambiguous
since its salience, frequency, experience and interpretation would
differ depending on the context, e.g., on whether it occurs during
physical exercise, a moment of anxiety, or quiet rest (e.g., someone who
exercises regularly, or is anxious, might more easily recall an instance
of occurrence of feeling their heart beating and respond based on that
contextual saliency).

@@

To further reduce and balance the remaining items, we collectively
decided on removing the lower-level clusters \emph{Sexual Arousal
Awareness} (too general and overlapping with the other more specific
sex-related items), \emph{Signalling} (which items started with ``when
something important is happening in my life'', which meaning we deemed
too open to interpretation), and \emph{Sexual Arousal Changes} (low fit
with the other modality-focused clusters of its group). The final set
included 45 items.

@@

Next, we tested the predictive power of the Mint scale relative to other
interoception questionnaires. We assessed the importance of each
interoceptive dimension (from all the scales) as a unique predictor by
fitting regression models (linear for continuous measures - e.g.,
depression score - and logistic for binary variables - e.g., presence
\emph{vs.} absence of mood disorders) to predict different outcomes, and
compare the best between the 4 interoception scales based on the total
explained variance (R2). We then evaluated the predictive performance of
each scale as a whole by comparing regression models with all the
dimensions entered as predictors (note that the IAS and BPQ only have
one total score variable). We assessed the models based on their R2, as
well as on the Bayesian Information Criterion (BIC), which penalizes for
predictor number, thus offering a balance between model parsimony and
predictive power. For the logistic models, we quantified the
discriminative power by computing the Area Under the Curve (AUC) of the
Receiver Operating Characteristic (ROC) curve, which assesses the
model's discriminative power (the combination of sensitivity and
specificity).

@@

The correlation matrix of the Mint dimensions revealed an interesting
and rich tapestry of relationships (see Figure~\ref{fig-four}), with
contrasting patterns of associations with dimensions from the same
group. For instance, \emph{Derm} and \emph{Olfa}, despite being
positively correlated with the other dimensions from the \emph{Deficit}
cluster, were not negatively correlated to \emph{Awareness} and its
dimensions. Similarly, \emph{Sati}, unlike the other \emph{Deficit}
dimensions, was not positively correlated to \emph{Visceroception}. This
complex structure suggests that the dimensions indeed capture distinct
aspects of interoception, and that the metaclusters, rather than being
simple aggregates of rather-redundant elements, might actually capture
unique combinations (greater than the sum of their parts). It also
provides evidence against a simplistic adaptive \emph{vs.} maladaptive
dichotomy, as \emph{Deficit} was not negatively but positively
correlated with \emph{Visceroception}. Interestingly, while
\emph{Awareness} was also positively correlated with
\emph{Visceroception}, it yielded an insignificant negative correlation
with \emph{Deficit} (this finding was also aligned with the hierarchical
clustering results, showing a greater proximity between \emph{Deficit}
and \emph{Visceroception} than with \emph{Awareness}), underlining again
the complex web of relationships captured by the Mint.
